---
output:
  github_document:
    html_preview: false
params:
  site_counts: NA
  global_grid: NA
always_allow_html: true
---

# Download process

```{r message=FALSE, echo=FALSE}
global_grid <- params$global_grid
site_counts <- params$site_counts

options(tigris_use_cache = TRUE)
```

### Catalogue existing data

The download process begins by cataloging the existing data that is available. We start out using the following parameters and parameter codes for retrieving data:

chlorophyll:

-   "Chlorophyll"
-   "Chlorophyll A"
-   "Chlorophyll a"
-   "Chlorophyll a (probe relative fluorescence)"
-   "Chlorophyll a (probe)"
-   "Chlorophyll a - Periphyton (attached)"
-   "Chlorophyll a - Phytoplankton (suspended)"
-   "Chlorophyll a, corrected for pheophytin"
-   "Chlorophyll a, free of pheophytin"
-   "Chlorophyll a, uncorrected for pheophytin"
-   "Chlorophyll b"
-   "Chlorophyll c"
-   "Chlorophyll/Pheophytin ratio"

secchi:

-   "Depth, Secchi disk depth"
-   "Depth, Secchi disk depth (choice list)"
-   "Secchi Reading Condition (choice list)"
-   "Secchi depth"
-   "Water transparency, Secchi disc"

doc:

-   "Organic carbon"
-   "Total carbon"
-   "Hydrophilic fraction of organic carbon"
-   "Non-purgeable Organic Carbon (NPOC)"

tss:

-   "Total suspended solids"
-   "Total Suspended Particulate Matter"

<br>

### Maps of data spread within the contiguous US:

Maps are presented below with counts of records across a grid. The grid is how records are grouped in download requests to the Water Quality Portal. Florida is mapped separately because of its high concentration of values. **Note:** 1) the top four grid cells in the FL map are included in both maps, 2) the counts here are for raw data that are not filtered for simultaneous records.

```{r}

# Combine counts in each grid_id with the grid polygons
grid_counts <- left_join(x = global_grid,
                         y = site_counts %>%
                           count(grid_id),
                         by = c("id" = "grid_id")) %>%
  st_transform(crs = 9311)

state_selection <- states(progress_bar = FALSE) %>%
  filter(!NAME %in% c("Alaska", "Hawaii", "American Samoa",
                      "Guam", "Puerto Rico",
                      "United States Virgin Islands",
                      "Commonwealth of the Northern Mariana Islands")) %>%
  st_transform(crs = 9311)

non_fl_map <- ggplot() +
  geom_sf(data = grid_counts %>%
            filter(!id %in% c(#10847:10850,
                              10668, 10669, 10670, 10489, 10490,
                              10491, 10309, 10310)),
          aes(fill = n)) +
  geom_sf(data = state_selection %>%
            filter(NAME != "Florida"), fill = NA, color = "white") +
  # geom_sf_text(data = grid_counts, aes(label = id)) +
  xlab(NULL) +
  ylab(NULL) +
  coord_sf(xlim = c(min(st_coordinates(state_selection)[,"X"]),
                    max(st_coordinates(state_selection)[,"X"])),
           ylim = c(min(st_coordinates(state_selection)[,"Y"]),
                    max(st_coordinates(state_selection)[,"Y"]))) +
  scale_fill_viridis_c("Record count") +
  theme_bw()

non_fl_map

fl_states <- state_selection %>%
  filter(NAME %in% c("Florida", "Georgia", "Alabama"))

fl_map <- ggplot() +
  geom_sf(data = grid_counts %>%
            filter(id %in% c(10847:10850, 10668, 10669, 10670, 10489, 10490,
                              10491, 10309, 10310)),
          aes(fill = n)) +
  geom_sf(data = fl_states, fill = NA, color = "black") +
  # geom_sf_label(data = grid_counts, aes(label = id)) +
  xlab(NULL) +
  ylab(NULL) +
  coord_sf(xlim = c(min(st_coordinates(fl_states)[,"X"]),
                    max(st_coordinates(fl_states)[,"X"])),
           ylim = c(min(st_coordinates(fl_states)[,"Y"]),
                    max(st_coordinates(fl_states)[,"Y"]))) +
  scale_fill_viridis_c("Record count") +
  theme_bw()

fl_map


```
