---
title: "DOC Cleaning"
author: "Michael Meyer"
date: "2023-02-01"
output: html_document
---
# Aggregating

```{r}
library(kableExtra)
library(feather)
library(tidyverse)
library(knitr)
library(lubridate)
library(forcats)
library(rvest)
library(scales)
library(ggthemes)

#COLOR-BLIND PALLETTE:#
#show_col(colorblind_pal()(8))


site_url <- 'https://help.waterdata.usgs.gov/parameter_cd?group_cd=%'

tables <- read_html(site_url) %>%
  html_nodes('table') %>%
  html_table()

pcodes <-tables[[1]] %>%
  janitor::clean_names() %>%
  mutate(parm_cd=str_pad(as.character(parameter_code), 5, pad = "0"))

raw_doc <- read_feather('../../wqp_data_aoi_formatted_filtered.feather') %>% 
  filter(parameter == "doc") %>%
  dplyr::select(date=ActivityStartDate,
                parameter=CharacteristicName,
                parm_cd=USGSPCode,
                units=ResultMeasure.MeasureUnitCode,
                SiteID=MonitoringLocationIdentifier,
                org=OrganizationFormalName,
                org_id=OrganizationIdentifier,
                time=ActivityStartTime.Time,
                value=ResultMeasureValue,
                sample_method=SampleCollectionMethod.MethodName,
                analytical_method=ResultAnalyticalMethod.MethodName,
                particle_size=ResultParticleSizeBasisText,
                date_time=ActivityStartDateTime,
                media=ActivityMediaName,
                type=ActivityMediaSubdivisionName,
                sample_depth=ActivityDepthHeightMeasure.MeasureValue,
                sample_depth_unit=ActivityDepthHeightMeasure.MeasureUnitCode,
                fraction=ResultSampleFractionText,
                status=ResultStatusIdentifier,
                field_comments=ActivityCommentText,
                lab_comments=ResultLaboratoryCommentText,
                result_comments=ResultCommentText) %>%
  left_join(pcodes, by='parm_cd') %>%
  mutate(year=year(date),
         units = trimws(units)) %>%
  filter(year >= 1984) %>%
  filter(media%in%c("Water","water")) %>%
  filter(type%in%c('Surface Water','Water','Estuary')) %>%
  rowid_to_column(.,"index")
```

# DATA LUMPING

As a first pass, I wanted to see how many analytical methods were represented
in the dataset.

```{r}
raw_doc %>%
  group_by(analytical_method) %>% 
  count() %>% ungroup() %>%
  mutate(prop = n/sum(n)) %>%
  arrange(desc(prop)) %>%
  ggplot() + 
  geom_histogram(aes(x = prop)) +
  ylab("Count") +
  xlab("Proportion of Analytical Methods") +
  theme_bw()
```

So, it looks like a few analytical methods are largely represented in the dataset, 
and the vast majority of differeing analytical methods only appear less than 2.5% of the time. 
The most consistently represented methods (i.e., those present in >= 2.5% of 
the total unique methods) are:

`raw_doc %>%
  group_by(analytical_method) %>% 
  count() %>% ungroup() %>%
  mutate(prop = n/sum(n)) %>%
  filter(prop >= .025) %>%
  arrange(desc(prop))%>%
  kable(.,'html',caption='Primary Analytical Methods') %>%
  kable_styling(position='center') %>%
  scroll_box(width='800px',height='500px')`


```{r}
aggregated_methods <- raw_doc %>%
  filter(!grepl("Oxygen|Nitrogen|Ammonia|Metals|E. coli|Anion|Cation|Phosphorus|Silica|PH|HARDNESS|Nutrient|Turbidity|Nitrate|Conductance|Alkalinity|Chlorophyll|Solids", x = analytical_method, ignore.case = TRUE)) %>%
  mutate(method_status = case_when(
    
    # DOC-specific:
    grepl("5310 B ~ Total Organic Carbon by Combustion-Infrared Method|Total Organic Carbon by Combustion|
          5310 B ~ Total Organic Carbon by High-Temperature Combustion Method|SM5310B|
          Organic-C, combustion-IR method|EPA 415.1|SM 5310 B|EPA 415.1M|TOC, combustion (SM5310B,COWSC)|
          DOC, combustion, NDIR (SM5310B)|TOC, combustion & CO2 detection|415.1|TOC, wu, combustion (5310B;DDEC)|
          SM185310B|DOC, wf, combustion (5310B;DDEC)|EPA Method 415.1 for Total Organic Carbon in aqueous matrices|
          SM 5310 B v20|DOC, 0.45u silver, persulfate IR",
          analytical_method,ignore.case=T) ~ "5310B + EPA 415.1 - Combustion + IR",
    
    grepl("5310 C ~ Total Organic Carbon in Water- Ultraviolet Oxidation Method|UV OR HEATED PERSULFATE OXIDATION|
          DOC, UV/persulfate (NYWSC; ALSC)|415.2|DOC, persulfate oxidation & IR|
          SM5310C|SM 5310 C|TOC, persulfate oxid (5310C; PA)|TOC, wu, persulfate (SM5310C;CO)|DOC, persulfate oxid, IR (COWSC)|
          Dissolved Organic Carbon in Water by Persulfate Oxidation and Infrared Spectrometry|TOC, persulfate-UV oxid (NYSDEC)|
          TOTAL ORGANIC CARBON (TOC) PERSULFATE-ULTRAVIOLET|TOC - Persulfate-Ultraviolet or Heated-Persulfate Oxidation Method|
          DOC, wu, persulfate (SM5310C;ND)|TOC, wu, persulfate (SM5310C;ND)|TOC, UV/persulfate/IR (USGS-NYL)|
          DOC, persulfate oxid (5310C; PA)|EPA 415.2|DOC, UV/persulfate (NYWSC; KECK)|
          5310 C ~ Total Organic Carbon by High-Temperature Combustion Method|415.2 M ~ Total Organic Carbon in Water|
          SM 5310 C, EPA 415.3|5310 C ~ Total organic carbon by Persulfate-UV or Heated-Persulfate Oxidation Method|
          DOC, wu, persulfate (SM5310C;CO)", 
          analytical_method,ignore.case=T) ~ "5310C + USGS O-1122-92 + EPA 415.2 - Persulfate-UV/Heated Persulfate Oxidation + IR",
    
    grepl("TOC, wet oxidation|WET OXIDATION METHOD|DOC,0.45um cap,acid,persulfateIR|
    5310 D ~ Total Organic Carbon in Water- Wet-Oxidation Method|DOC, wf, 0.45 um cap, combust IR|415.3|
    Determination of Total Organic Carbon and Specific UV Absorbance at 254 nm in Source Water and Drinking Water|
    EPA 415.3|SM 5310 D|O-3100 ~ Total Organic Carbon in Water", 
          analytical_method,ignore.case=T) ~ "5310D + EPA 415.3 + USGS O-3100 - Wet Oxidation + Persulfate + IR",
    
    grepl("440 W  ~ Determination of Carbon and Nitrogen", 
          analytical_method,ignore.case=T) ~ "EPA 440.0",
    
    grepl("9060 A ~ Total Organic Carbon in water and wastes by Carbonaceous Analyzer|9060 AM ~ Total Volatile Organic Carbon|
          EPA 9060|EPA 9060A",
          analytical_method,ignore.case=T) ~ "EPA 9060A - Carbonaceous Analyzer",
    !grepl("5310 B ~ Total Organic Carbon by Combustion-Infrared Method|Total Organic Carbon by Combustion|
          5310 B ~ Total Organic Carbon by High-Temperature Combustion Method|SM5310B|
          Organic-C, combustion-IR method|EPA 415.1|SM 5310 B|EPA 415.1M|TOC, combustion (SM5310B,COWSC)|
          DOC, combustion, NDIR (SM5310B)|TOC, combustion & CO2 detection|415.1|TOC, wu, combustion (5310B;DDEC)|
          SM185310B|DOC, wf, combustion (5310B;DDEC)|EPA Method 415.1 for Total Organic Carbon in aqueous matrices|
          SM 5310 B v20|DOC, 0.45u silver, persulfate IR|5310 C ~ Total Organic Carbon in Water- Ultraviolet Oxidation Method|
          UV OR HEATED PERSULFATE OXIDATION|
          DOC, UV/persulfate (NYWSC; ALSC)|415.2|DOC, persulfate oxidation & IR|
          SM5310C|SM 5310 C|TOC, persulfate oxid (5310C; PA)|TOC, wu, persulfate (SM5310C;CO)|DOC, persulfate oxid, IR (COWSC)|
          Dissolved Organic Carbon in Water by Persulfate Oxidation and Infrared Spectrometry|TOC, persulfate-UV oxid (NYSDEC)|
          TOTAL ORGANIC CARBON (TOC) PERSULFATE-ULTRAVIOLET|TOC - Persulfate-Ultraviolet or Heated-Persulfate Oxidation Method|
          DOC, wu, persulfate (SM5310C;ND)|TOC, wu, persulfate (SM5310C;ND)|TOC, UV/persulfate/IR (USGS-NYL)|
          DOC, persulfate oxid (5310C; PA)|EPA 415.2|DOC, UV/persulfate (NYWSC; KECK)|
          5310 C ~ Total Organic Carbon by High-Temperature Combustion Method|415.2 M ~ Total Organic Carbon in Water|
          SM 5310 C, EPA 415.3|5310 C ~ Total organic carbon by Persulfate-UV or Heated-Persulfate Oxidation Method|
          DOC, wu, persulfate (SM5310C;CO)|TOC, wet oxidation|WET OXIDATION METHOD|DOC,0.45um cap,acid,persulfateIR|
    5310 D ~ Total Organic Carbon in Water- Wet-Oxidation Method|DOC, wf, 0.45 um cap, combust IR|415.3|
    Determination of Total Organic Carbon and Specific UV Absorbance at 254 nm in Source Water and Drinking Water|
    EPA 415.3|SM 5310 D|O-3100 ~ Total Organic Carbon in Water|9060 A ~ Total Organic Carbon in water and wastes by Carbonaceous Analyzer|9060 AM ~ Total Volatile Organic Carbon|
          EPA 9060|EPA 9060A",
          analytical_method,ignore.case=T) & !is.na(analytical_method) ~ "Ambiguous"))

grouped_more <- aggregated_methods %>% 
  mutate(grouped = case_when(grepl(pattern = "5310B", 
                                  x = method_status) ~ "Combustion+IR",
                            grepl(pattern = "5310C", 
                                  x = method_status) ~ "Persulfate-UV/Heated Persulfate Oxidation+IR",
                            grepl(pattern = "5310D", 
                                  x = method_status) ~ "Wet Oxidation+Persulfate+IR",
                            grepl(pattern = "EPA 440.0", 
                                  x = method_status) ~ "Elemental Analyzer",
                            grepl(pattern = "EPA 9060A", 
                                  x = method_status) ~ "Carbonaceous Analyzer",
                            method_status == "Ambiguous" ~ "Ambiguous")) %>%
  mutate(aquasat_fraction=case_when(fraction %in% c('Dissolved','Filtered, lab','Filterable', 
                                                    'Filtered, field') ~ "Dissolved",
                                    fraction %in% c('Dissolved','Filtered, lab','Filterable') ~ "Dissolved",
                                    fraction %in% c('Total','Total Recovrble',
                                                    'Total Recoverable','Recoverable','Unfiltered',
                                                    "Acid Soluble", "Suspended", "Non-Filterable (Particle)") ~ "Total",
                                    fraction %in% c('Fixed') ~ "Fixed",
                                    fraction %in% c('Non-Filterable (Particle)') ~ 'Particle',
                                    is.na(fraction)|fraction%in%c(" ","Field", "Bed Sediment",
                                                                  "Inorganic", "Organic") ~ "Ambiguous"))

rm(site_url,tables,aggregated_methods,pcodes)
```

# Methods

In the water quality portal, DOC had `r n_distinct(raw_doc$analytical_method)` unique analytical methods listed for a total of `r nrow(grouped_more)` samples. However, many of these analytical methods can be grouped together into just a few methodologies.

These methods broadly represent the diversity of methods that are included 
within the [National Environmental Methods Index (NEMI)](https://www.nemi.gov/home/).

1. *Standard Method 5310B and EPA 415.1*: Combustion of sample followed by measurement 
of carbon with a nondispersive infrared analyzer. 

2. *Standard Method 5310C, USGS O-1122-92, EPA 415.2*: Oxidation of the sample
either by persulfate in the presence of heat or ultraviolet light, followed by 
the measurement of carbon with a nondispersive infrared analyzer.

3. *Standard Method 5310D 5310D, USGS O-3100, and EPA 415.3*: The sample is first
treated with an acid, and then oxidized with persulfate in the presence of heat 
or ultraviolet light. Following oxidation, carbon is measured using a nondispersive infrared analyzer.

4. *EPA 440.0*: This methods is largely for particulate organic carbon and nitrogen. 
A massed amount of dried particulate or sediment is combinsted at a high temperature 
using an elemental analyzer. 

```{r}
grouped_more %>%
  group_by(grouped) %>%
  summarize(count = n()) %>%
  arrange(count) %>%
  ggplot()+
  geom_col(aes(x = fct_reorder(grouped,count),y = count, fill = grouped)) +
  geom_text(aes(label = count, y = count/2, x = fct_reorder(grouped,count), 
                color = grouped),
                position = position_dodge(0.5),
                vjust = 0) +
  ylab("Count") +
  xlab("") +
  coord_flip() +
  theme_bw() +
  scale_fill_manual(values=c("#CC79A7","#E69F00","#56B4E9","#009E73","#0072B2")) + 
  scale_color_manual(values=c("black","black","black","black","black","black")) + 
  scale_y_continuous(labels = comma)+
  theme(legend.position = "none",
        text = element_text(size = 20))
```


Methods using nondispersive infrared analyzers are the most common across grouped methods (at r grouped_more %>% mutate(total=n()) %>% group_by(grouped) %>% filter(grepl("IR", grouped)) %>% ungroup() %>% summarize(perc=(n()/total)*100) %>% distinct() %>% ungroup() %>% select(perc) %>% round(digits=3) %>% paste()%).

However, there are differences in how a sample can be treated prior to analysis on the nondispersive infrared analyzer. 

```{r}
pie <- grouped_more %>%
  filter(grepl(pattern = "IR", x = grouped)) %>%
  group_by(method_status) %>%
  summarize(count=n()) %>%
  ungroup() %>%
  mutate(method_status=ifelse(is.na(method_status), "NDIR", method_status))
pie <- pie %>% mutate(method_status = factor(x = method_status, levels = method_status)) %>% 
  mutate(prop = count/sum(pie$count)) %>%  
  mutate(ypos = cumsum(prop)- 0.5*prop) %>%
  mutate(legend = paste0(method_status, " (", percent(prop), ")"))

ggplot(data=pie, aes(x="", y=count, fill=legend)) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) +
  scale_fill_manual(values=c("#009E73","#E69F00","#56B4E9","#CC79A7","#0072B2","#F0E442","#D55E00")) +
  guides(fill=guide_legend(title="NDIR Methods")) +
  theme_void() + # remove background, grid, numeric label
  theme(text = element_text(size = 20))
```

# Sample Fractions

When looking at relative fractions between methodologies, we see that the proportion 
of total and dissolved fractions are approximately the same across all methods. 

```{r}
grouped_more %>%
  group_by(grouped, aquasat_fraction) %>%
  summarize(count=n()) %>%
  ggplot(aes(x=grouped,y=count,fill=aquasat_fraction)) +
           geom_bar(stat="identity", width=1, color="white") +
  theme_bw() +
  theme(text = element_text(size = 15)) +
  coord_flip() +
  ylab("Grouped Methodologies")+
  xlab("") +
  scale_fill_manual(values=c("#CC79A7","#0072B2","#F0E442","#D55E00","#009E73")) +
  guides(fill=guide_legend(title="Sample Fraction")) +
  scale_y_continuous(labels = comma)
```

# How to tier DOC

There is no clear way of tiering silica based on fraction if we want all tiers to be looking at the exact same thing. Instead, I suggest we only look at samples that are analyzing the dissolved fraction. The dissolved fraction makes up r grouped_more %>% mutate(total=n()) %>% group_by(aquasat_fraction) %>% filter(aquasat_fraction=="Dissolved") %>% summarize(perc=(n()/total)*100) %>% distinct() %>% ungroup() %>% select(perc) %>% round(digits=3) %>% paste()% of all silica samples in the Water Quality Portal, a clear sign that the majority of people monitoring silica are looking at the dissolved fraction. With this in mind, I suggest tiering silica as follows:

Restrictive. ICP (all versions). Dissolved fraction. Only water samples with an accepted value that is reasonable with logical units.

Narrowed. ICP and colorimetry (all versions). Dissolved fraction. Only water samples with an accepted value that is reasonable with logical units.

Inclusive. ICP and colorimetry (all versions), and all other non-ambiguous methods. Dissolved fraction. Only water samples with an accepted value that is reasonable with logical units. Currently no samples fall within this tier.

Ambiguous (and therefore removed). Ambiguous methods. Dissolved fraction. Only water samples with an accepted value that is reasonable with logical units.

Without any additional steps other than harmonizing units and removing NA values (I haven't looked at whether values are reasonable), this is how that data looks:

```{r}
#identify samples that have no meaningful data related to an NA value
no_data_samples <- grouped_more %>%
  filter(is.na(value)&is.na(units)&is.na(lab_comments)&is.na(result_comments)) 

#remove samples that have no values and no lab/result metadata
doc_empties_removed <- grouped_more %>%
  filter(status %in% c('Accepted','Final','Historical','Validated')) %>%
  filter(!index %in% no_data_samples$index)

vals_cleaned <- doc_empties_removed %>%
  mutate(numeric_value=as.numeric(value)) %>%
  filter(!is.na(numeric_value)) # need to work on this

# unit_disharmony <- function(d,lookup){
#   d %>%
#     anti_join(silica_lookup,by='units') %>%
#     group_by(units) %>%
#     summarize(count=n())  %>%
#     kable(.,'html',caption='The following measurements
#           were dropped because the units do not make sense') %>%
#     kable_styling() %>%
#     scroll_box(width='500px',height='400px')
# }

#Set up a lookup table so that final units are all in ug/L. 
doc_lookup <- tibble(units=c('mg/L','mg/l','ppm','ug/l','ug/L','mg/m3','ppb','mg/cm3','ug/ml','mg/ml','ppt','umol/L'),
                        conversion = c(1000,1000,1000,1,1,1,1,1000000,1000,1000000,0.000001,60.080000))

#unit_disharmony(vals_cleaned,silica_lookup)

doc_harmonized <- vals_cleaned %>%
  inner_join(doc_lookup,by='units') %>%
         mutate(harmonized_value=(numeric_value*conversion)/1000,
         harmonized_unit='mg/L')

doc_tiered <- doc_harmonized %>%
  filter(aquasat_fraction=="Dissolved") %>%
  mutate(tiers=case_when(grouped %in% c("Wet Oxidation+Persulfate+IR",
                                        "Persulfate-UV/Heated Persulfate Oxidation+IR") ~ "Restrictive",
                         grouped=="Combustion+IR" ~ "Narrowed",
                         grouped %in% c("Ambiguous", "Carbonaceous Analyzer") ~ "Inclusive",
                         is.na(grouped) ~ "Dropped from Aquasat")) 


# meanscores <- attributes(silica_tiered$grouped)$harmonized_value
# 
# meandf <- data.frame(
#   variable = rep(names(meanscores), 4),
#   value    = rep(unname(meanscores), 6),
#   cluster  = rep(1:6, each=14)
#   )

ggplot(data=doc_tiered) +
  geom_histogram(aes(x=(harmonized_value), fill=tiers), bins=100) +
  scale_x_log10(breaks = trans_breaks("log10", function(x) 10^x),
                labels = trans_format("log10", math_format(10^.x))) +
  facet_wrap(~tiers, scales='fixed') +
  xlab('DOC mg/L') +
  theme_bw() +
  theme(legend.position = "none",
        text = element_text(size = 20))
```

Remaining questions

Is this tiering appropriate? Does not take into account differences in which method is better for a specific range of values. I'm also lumping all ICP methods (ICP-MS, ICP-AES, ICP, etc.) together.

This approach to tiering tosses out what could be perfectly good data (e.g. high-quality "total" silica). Perhaps not an issue for silica, but this could be a pretty subjective way to tier other metals when we get to them?

How to handle non-detects/BDLs. In the upstream workflow I am removing all samples with non-numeric values. However, many samples have information within the value field that could be used to determine samples that were below the detection limit:

```{r}
doc_empties_removed %>%
  mutate(numeric_value=as.numeric(value)) %>%
  filter(is.na(numeric_value)) %>%
  distinct(value,.keep_all = T) %>%
  filter(!is.na(value)) %>%
  arrange(desc(value)) %>%
  select(value) %>%
  kable(.,'html',caption='Text Inputted in Value Field') %>%
  kable_styling() %>%
  scroll_box(width='900px',height='400px')
```  

There are also a lot of samples that are shown to be UNDER some value (i.e. < 2.6). For these, I've played around with changing their value to a random number between 0 and half of the listed value, though this definitely makes some pretty large/maybe inappropriate assumptions about what these values actually mean! I've flagged these in the `aquasat_comments` field as being *Approximated, EPA MDL method*.

```{r}
mdl <- doc_empties_removed %>%
  select(index,value) %>%
  filter(grepl("0|1|2|3|4|5|6|7|8|9",value) & grepl("<", value))

mdl$num_value <- as.numeric(str_replace_all(mdl$value, c("\\<"="", "\\*" = "")))

mdl <- mdl %>%
  mutate(zero = 0,
         half = num_value/2)

mdl$epa_mdl <- with(mdl,runif(nrow(mdl),zero,half))

mdl <- select(mdl,index,epa_mdl)

# replace value field with these new values
mdls_added <- doc_empties_removed %>%
              left_join(mdl,by="index") %>%
              mutate(aquasat_value=ifelse(index %in% mdl$index, epa_mdl, value),
                     aquasat_comments = ifelse(index %in% mdl$index, "Approximated, EPA LDL method", NA))

approx <- mdls_added %>%
  filter(!index %in% mdl$index) %>% # remove the samples that we've already approximated using the EPA method
  mutate(num_value = as.numeric(value)) %>%
  filter(is.na(num_value) & 
           grepl("0|1|2|3|4|5|6|7|8|9", value) &
           (grepl("result approx|RESULT IS APPROX", lab_comments, ignore.case=T)|
            grepl("result approx|RESULT IS APPROX", result_comments, ignore.case=T ))) #select samples that are non numeric but have a number listed as well as comments related to approximation

approx$approx_value <- as.numeric(str_replace_all(approx$value, c("\\*" = "")))
approx <- select(approx,index,approx_value)

# replace value field with these new values

approx_added <- mdls_added %>%

              left_join(approx,by="index") %>%

              mutate(aquasat_value=ifelse(index %in% approx$index, approx_value, aquasat_value)) %>%

              mutate(aquasat_comments=ifelse(index %in% approx$index, "Approximate", aquasat_comments))
```