---
output: html_document
editor_options: 
  chunk_output_type: console
---


# (PART) Data downloads and decisions {-}

# Download and Join


I recognize that I should probably just use `targets` for this part, but...
I (Matt Ross) don't know how to use targets yet! So, instead we are just
going to have several code chunks here that just downloads all the data directly
from figshare, preps data for analysis, etc...
and the chunks will just be set to `eval = FALSE`, so they won't
run unless you manually run it. That way this book can run in reasonable timeframes,
most other chunks will run based on "caching." Though, some data intensive tasks,
will also simply have `eval = F` to save runtime. These chunks will be flagged
explicitly, and all their data products saved to a `data/out` folder. 

## AquaSat download

AquaSat is held in a figshare collection with many pieces. For this work we
will need most of the data held there, which we will access through direct download.
The URLs are just from the AquaSat collection links.

```{r, eval = F}
## AquaSat Download data. This takes > 2 hours,
## so only run if you think 
## AquaSat has been updated!

## Also

## In-situ raw data with methods
download.file('https://figshare.com/ndownloader/files/15475154',
              mode = 'wb',# Needs to be written in binary for some reason 
              destfile = 'data/in/aq_situ.zip')
unzip('data/in/aq_situ.zip',exdir = 'data/in/aq_situ')


# Site Inventory with type, because it's not in the other inventory
# Stupid aquasat developer (me/mattross!)
download.file('https://figshare.com/ndownloader/files/24720434',
              mode = 'wb',
              destfile = 'data/in/inv.feather')


## Unique site inventory 

download.file('https://figshare.com/ndownloader/files/24720437',
              mode = 'wb',
              destfile = 'data/in/unq_site.feather')

## Ecoregion data
download.file('https://gaftp.epa.gov/EPADataCommons/ORD/Ecoregions/cec_na/na_cec_eco_l2.zip', destfile = 'data/in/eco2.zip')

unzip('data/in/eco2.zip',exdir = 'data/in/ecoregion')


```

## AquaSat Read and join. 

Note, also eval = F to save computation time

```{r, eval = F}

library(tidyverse)
library(sf)
library(feather)
library(GGally)
library(mapview)

# This is only useful because it has type (lake/stream/etc...)
inv_type <- read_feather('data/in/inv.feather') %>%
    select(SiteID = MonitoringLocationIdentifier,
         type = ResolvedMonitoringLocationTypeName) %>%
  mutate(type = ifelse(grepl('Lake',type),'Lake',type))


#Real inventory
site.vis <- read_feather('data/in/unq_site.feather') %>%
  inner_join(inv_type) %>%
  distinct(SiteID,lat,long,type)



# Complete in-situ data, 6.5 million records. 
in.vis <- read_csv('data/in/aq_situ/in-situ/wqp_lagos_unity.csv') 

# Get clouds and path row sites

clouds <- read_feather('data/in/clouds.feather') %>%
  mutate(date = as.Date(SENSING_TIME)) %>%
  select(PATH = WRS_PATH, 
         ROW = WRS_ROW,
         clouds = CLOUD_COVER,
         date)


site_path_row <- site.vis %>%
  st_as_sf(.,coords = c('long','lat'), crs = 4326) %>%
  st_join(.,st_read('data/in/wrs/WRS2_descending.shp')) %>%
  select(SiteID, PATH, ROW) %>%
  as_tibble(.) %>%
  select(-geometry)

```

## Data Eval and Subset


### Selecting only simultaneous observations

I think the code mostly speaks for itself here
```{r, eval = F}
# Only bother with data that has complete simultaneous observations of
# chl_a, doc, etc...

simul_vis <- in.vis %>%
  select(-p_sand) %>%
  filter(if_all(c(chl_a,doc,tss,secchi), ~!is.na(.))) %>% 
  inner_join(site.vis) %>%
  filter(type != 'Facility') %>%
  #Set some reasonable thresholds, AquaSat is too generous
  filter(secchi < 15,
         chl_a < 1000,## ug/L
         tss < 1000, ## mg/L
         doc < 50)

no_secchi <- in.vis %>%
  select(-p_sand) %>%
  filter(across(c(chl_a,doc,tss), ~!is.na(.))) %>% 
  inner_join(site.vis) %>%
  filter(type != 'Facility') %>%
  #Set some reasonable thresholds, AquaSat is too generous
  filter(
         chl_a < 1000,## ug/L
         tss < 1000, ## mg/L
         doc < 50)

# For speed
write_feather(simul_vis, 'data/out/simul.feather')
#For reproducibility
write_csv(simul_vis, 'data/out/simul.csv')

#For speed
write_feather(no_secchi, 'data/out/no_secchi.feather')



```



## Where are sites with simultaneous observations of clarity constituents? 


```{r}

simul_vis <- read_csv('data/out/simul.csv')

unique_simul <- simul_vis %>%
  distinct(SiteID, lat, long, type) %>%
  st_as_sf(.,coords = c('long','lat'), crs = 4326)

mapviewOptions(fgb = FALSE) # remove flat geobuff which breaks display
mapview(unique_simul, zcol = 'type')

```


## What is the general relationiship between variables in log log space

```{r, cache = T}


log_simul <- simul_vis %>%
  dplyr::mutate(across(c(secchi,chl_a,tss,doc,tis),log10)) %>%
  dplyr::filter(across(c(chl_a,doc,secchi,tss), ~!is.na(.) & . < Inf & . > -Inf)) 


log_simul %>%
  sample_frac(0.2) %>%
  ungroup() %>%
  select(secchi,chl_a,tss,doc,type) %>%
  ggpairs(lower = list(continuous = wrap('points',shape = 1)),
          diag = list(continuous = wrap('densityDiag', alpha = 0.5)),
         mapping = ggplot2::aes(color = type),
         columns = c('secchi','chl_a','tss','doc')) +
  ggthemes::theme_few() + 
  scale_color_manual(values = c('seagreen3','skyblue3','saddlebrown'))

```


